---
title: "{{< var maxyr >}} Results Summary"
params:
  maxyr: 2025
bibliography: refs.bib
---

```{r}
library(tbeptools)
library(glue)
library(tidyverse)
library(patchwork)
library(here)
library(flextable)
library(ftExtra)
library(ggspatial)
library(ggrepel)
library(sf)

source('R/funcs.R')

data(chldat)

# for matrix and ra tables, must conform to tbeptools format
chldatmod <- chldat %>%
  rename(
    epchc_station = station
  ) %>%
  mutate(
    sd_m = 1,
    bay_segment = case_when(
      bay_segment %in% c('BCBS', 'TCB', 'MR') ~ 'RALTB',
      T ~ bay_segment
    )
  )
```

Results from {{< var maxyr >}} indicate that all RA bay segments met chlorophyll-a thresholds accepted by the FDEP to maintain FDEP Reasonable Assurance for Tampa Bay and to comply with the EPA TMDL (@fig-thrplot) and estuarine numeric nutrient criteria for Tampa Bay ([EPA Amended Approval Letter Jun. 28, 2013](https://www.epa.gov/sites/default/files/2015-04/documents/florida-amended-determination.pdf)). Chlorophyll-a concentrations were elevated throughout the bay during late summer/fall months in {{< var maxyr >}} due to the influence of several tropical storms on the region (@fig-boxplot). Mixed-assemblage algae blooms were reported in Hillsborough Bay, Old Tampa Bay, and Middle Tampa Bay, while low concentrations of *Karenia brevis* continued to be observed in Lower Tampa Bay into the winter months. Although understanding and mitigating blooms of *Pyrodinium bahamense* in OTB continues to be a focus of research and management efforts, summer concentrations in Old Tampa Bay were lower compared to prior years and the chlorophyll-a criteria was met for the third consecutive year of the RA period.  Preliminary recommendations from the assimilative capacity assessment for Old Tampa Bay to evaluate the existing management paradigm and assess the potential need to adopt alternative indicators and/or load allocations to address recurring water quality issues in that bay segment are underway [@tbep0924;@tbep0525].

The TBEP, in partnership with the Southwest Florida Water Management District, has previously developed an integrated ecosystem model to evaluate the net environmental benefits that may result from implementing various management actions in Old Tampa Bay including: reducing point sources, nonpoint sources, and causeway obstructions in Old Tampa Bay [@tbep1015]. Management actions that proximate and respond to current Old Tampa Bay conditions will be further evaluated under the Old Tampa Bay assimilative capacity assessment project. Furthermore, the TBEP funded research conducted by the Florida Fish and Wildlife Research Institute to improve understanding of the cell physiology and behavior of *Pyrodinium bahamense* and evaluate the potential for using shellfish to mitigate these algal blooms in Old Tampa Bay.  A water quality dashboard ([https://shiny.tbep.org/wq-dash](https://shiny.tbep.org/wq-dash)) continues to be available to further synthesize available data, assess additional water quality metrics (phytoplankton counts), and inform Consortium participants and other resource managers on the status of water quality in Tampa Bay. The dashboard allows for proactive response to anomalous water quality conditions on a month-to-month basis by the community.  As part of the Old Tampa Bay assimilative capacity assessment, potential modifications to existing modeling tools and the application of additional models will be finalized in 2025.

Between 2022 and 2024, seagrasses throughout Tampa Bay increased by 1,407 acres. After several consecutive years of losses, aerial photographs taken in December 2023 - February 2024 indicate that Tampa Bay now harbors 31,544 acres of seagrass. Seagrass acreage showed the greatest increases in Hillsborough Bay (+756 acres), accounting for more than half of the increase observed baywide. Despite modest gains elsewhere, Old Tampa Bay lost an additional 326 acres of seagrass, the lowest estimate ever recorded for this bay segment.  These trends are generally corroborated by annual transect surveys conducted by TBEP and its partners (shown in the [seagrass transect dashboard](https://shiny.tbep.org/seagrasstransect-dash)). Additional research and discussion is being pursued by the [Southwest Florida Seagrass Working Group](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#seagrass) to better understand the underlying mechanisms influencing these observations. Analysis of the role of long-term temperature increases and salinity reductions related to climate change and the potential relationships with seagrass declines since 2016 was published in 2024 [@tbep1124]. The next SWFWMD seagrass coverage estimate will be developed from aerial photographs acquired over the winter 2025-26 period. 

Detailed results for the 2022-2026 RA implementation period are provided in Tables [-@tbl-raotb], [-@tbl-rahb], [-@tbl-ramtb], [-@tbl-raltb], and [-@tbl-raraltb] for each bay segment. Results for the Remainder Lower Tampa Bay segment (Boca Ciega Bay South, Terra Ceia Bay, Manatee River) are also included in this annual assessment (reporting began with the 2023 annual assessment). As of the {{< var maxyr >}} reporting period, NMC Actions 2-5 are not necessary based upon observed water quality conditions within Tampa Bay, though additional work is being pursued by the TBEP and TBNMC to understand the most recent trends in seagrass coverage and Old Tampa Bay's current assimilative capacity. Individual annual reports of the bay’s conditions from {{< var maxyr >}} can be found on the TBEP website, as specified in the following link [@tbep0325] and the [water quality dashboard](https://shiny.tbep.org/wq-dash). A summary of historic attainment of the regulatory chlorophyll-a thresholds for each of the bay segments is depicted in @fig-chlmat.

Lastly, annual hydrologic conditions among all the major bay segments in {{< var maxyr >}} were above 1992-1994 levels ([@tbl-hydrotab]). Therefore, hydrologic adjustments for evaluating compliance with individual entity load allocations/permitting targets should be applied for each major segment [@tbep0512;@tbep0316]. The estimated hydrologic loads for each bay segment relative to observed 1992-1994 levels are indicated in the table below. The estimated compliance load adjustment factors (if applicable) are also specified. A tool to calculate the hydrologic estimates and adjustment factors by bay segment is available online through an interactive dashboard and automatically updated as provisional hydrologic estimates are approved by monitoring agencies (<https://shiny.tbep.org/tbnmc_hydrologic_estimates/>).

```{r}
#| label: tbl-hydrotab
#| tbl-cap: !expr paste0("Hydrologic load estimates in ", params$maxyr, " relative to 1992-1994 levels and estimated compliance load adjustment factors for the major bay segments.")
hydrotab(params$maxyr, Sys.getenv('NOAA_KEY'))
```

```{r}
#| label: fig-thrplot
#| fig-cap: !expr paste0("Historic chlorophyll-a annual averages for the four major bay segments of Tampa Bay and those that include the Remainder Lower Tampa Bay segment (Boca Ciega Bay South, Terra Ceia Bay, Manatee River). Annual averages in ", params$maxyr, " were below the regulatory thresholds developed under the Tampa Nitrogen Management Consortium’s nutrient management strategy in all bay segments. Vertical grey bars indicate the portion of the 2022-2026 Reasonable Assurance compliance assessment period covered by the results.")
#| fig-align: center
#| fig-width: 12
#| fig-height: 14
show_rachlplot(chldat, params$maxyr)
```

```{r}
#| label: fig-sitemap
#| fig-cap: !expr paste0("Map depicting individual station chlorophyll-a exceedences in Tampa Bay relative to FDEP regulatory thresholds for chlorophyll-a in ", params$maxyr, ". *Note individual station exceedences do not indicate failed compliance at the bay segment scale.*")
#| fig-align: center
#| fig-height: 8
#| fig-width: 5
#| out-width: "70%"
show_rasitemap(chldat, params$maxyr, thrs = T)
```

```{r}
#| label: fig-boxplot
#| fig-cap: !expr paste0(params$maxyr, " monthly chlorophyll-a bay segment means (red dots) compared to monthly distributions from prior years (box plots and black dots).  Prior years extend to 1975 for Old Tampa Bay, Hillsborough Bay, Middle Tampa Bay, and Lower Tampa Bay, 1991 for Boca Ciega Bay South, 1989 for Terra Ceia Bay, and 1990 for Manatee River. Note that Pinellas and Manatee County data are used for Boca Ciega Bay South, Terra Ceia Bay, and Manatee River and has less frequent sampling intervals than data from the Environmental Protection Commission of Hillsborough County used for the other bay segments. Boxes encompass the 25th and 75th percentiles, while whiskers bound the interquartile range. Dots beyond the whiskers represent outliers throughout the 1975-", params$maxyr -1, " sample period.")
#| fig-height: 15
#| fig-width: 12
#| fig-align: center
show_chlboxplot(chldat, params$maxyr)
```

```{r}
#| label: fig-seagrass
#| fig-cap: "Historic seagrass coverage estimates for Tampa Bay. The target coverage of 38,000 acres was changed to 40,000 acres in 2020 to reflect programmatic goals in the 2020 Habitat Master Plan Update ([TBEP #07-20](https://drive.google.com/file/d/1Hp0l_qtbxp1JxKJoGatdyuANSzQrpL0I/view?usp=drivesdk)). Data source: TBEP & SWFWMD."
#| fig-height: 3.25
#| fig-width: 6
#| fig-align: center
#| out-width: "90%"
show_seagrasscoverage(seagrass, maxyr = 2024, lastlab = T)
```

```{r}
#| label: fig-chlmat
#| fig-cap: !expr paste0("Attainment of adopted chlorophyll-a thresholds (1975 - ", params$maxyr, ") in the four major bay segments and Remainder Lower Tampa Bay segment (Boca Ciega Bay South, Terra Ceia Bay, Manatee River).  Green (yes) indicates that average annual chlorophyll-a thresholds were met; red (no) indicates that threshold levels were not met.  Grey line is the beginning of the current Reasonable Assurance implementation period.")
#| fig-height: 10
#| fig-width: 4
#| fig-align: center
#| out-width: "40%"
show_chlmatrix(chldatmod, params$maxyr)
```

```{r}
#| label: tbl-raotb
#| tbl-cap: !expr ratabcap('OTB')
show_ratab(bay_segment = 'OTB', yrsel = params$maxyr, epcdata = chldatmod)
```

```{r}
#| label: tbl-rahb
#| tbl-cap: !expr ratabcap('HB')
show_ratab(bay_segment = 'HB', yrsel = params$maxyr, epcdata = chldatmod)
```

```{r}
#| label: tbl-ramtb
#| tbl-cap: !expr ratabcap('MTB')
show_ratab(bay_segment = 'MTB', yrsel = params$maxyr, epcdata = chldatmod)
```

```{r}
#| label: tbl-raltb
#| tbl-cap: !expr ratabcap('LTB')
show_ratab(bay_segment = 'LTB', yrsel = params$maxyr, epcdata = chldatmod)
```

```{r}
#| label: tbl-raraltb
#| tbl-cap: !expr ratabcap('RALTB')
show_ratab(bay_segment = 'LTB', yrsel = params$maxyr, epcdata = chldatmod) # same as ltb for ra
```
